---
title: "Homework 2"
author: "Christine Lucille Kuryla (clk2162)"
date: "2023-10-03"
output: github_document
---

## Problem 1

```{r setup}
library(tidyverse)
```

Import and clean data from fivethirtyeight datasets of interest.

```{r q1_fivethirtyeight}

# Import and clean data from fivethirtyeight datasets pols-month

pols_df <- read_csv("./data/fivethirtyeight_datasets/pols-month.csv") %>% 
  separate(mon, into = c("year", "month", "day"), sep = "-") %>% 
  mutate(month = 
           case_match(
             month,
             "01" ~ "january",
             "02" ~ "february",
             "03" ~ "march",
             "04" ~ "april",
             "05" ~ "may",
             "06" ~ "june",
             "07" ~ "july",
             "08" ~ "august",
             "09" ~ "september",
             "10" ~ "october",
             "11" ~ "november",
             "12" ~ "december",
           )) %>%
  mutate(president = case_when(
    prez_dem == 1 ~ "dem",
    prez_dem == 0 ~ "rep"
      )) %>% 
  select(-c("prez_dem", "prez_gop", "day"))


# note that the mutate method for prez_dem == 1 or == 0 was chosen because of the code book, and because there exists values of "2" in prez_gop, which does not agree with the code book. Here we are assuming that prez_rep = 2 and prez_dem = 0 means that the president was republican. 
  

# Import and clean snp.csv 

snp_df <- read_csv("./data/fivethirtyeight_datasets/snp.csv") %>% 
    separate(date, into = c("month", "day", "year"), sep = "/") %>% 
  mutate(month = 
           case_match(
             month,
             "1" ~ "january",
             "2" ~ "february",
             "3" ~ "march",
             "4" ~ "april",
             "5" ~ "may",
             "6" ~ "june",
             "7" ~ "july",
             "8" ~ "august",
             "9" ~ "september",
             "10" ~ "october",
             "11" ~ "november",
             "12" ~ "december",
           )) %>%
  mutate(year = paste0("20",year)) %>%
  relocate(c("year", "month")) %>%
  select(-"day")

# import and tidy unemployment.csv

unemployment_df <- read_csv("./data/fivethirtyeight_datasets/unemployment.csv") %>% 
  janitor::clean_names() %>% 
  pivot_longer(
    "jan":"dec",
    names_to = "month",
    values_to = "unemployment"
  ) %>% 
    mutate(month = 
           case_match(
             month,
             "jan" ~ "january",
             "feb" ~ "february",
             "mar" ~ "march",
             "apr" ~ "april",
             "may" ~ "may",
             "jun" ~ "june",
             "jul" ~ "july",
             "aug" ~ "august",
             "sep" ~ "september",
             "oct" ~ "october",
             "nov" ~ "november",
             "dec" ~ "december",
           )) %>% 
  mutate(year = as.character(year)) # this is to enable the merge in the next step

```

Join the datasets by mergine snp into pols, and merging unemployment into the result.

```{r q1_merge}

# Merge snp_df into pols_df

pols_snp_df <- left_join(pols_df, snp_df)

# Merge unemployment data in as well

pols_snp_unemployment_df <- left_join(pols_snp_df, unemployment_df)

pols_snp_unemployment_df <- pols_snp_unemployment_df %>%
  mutate(year = as.numeric(year))

```

The dataset contains the following variables: `r colnames(pols_snp_unemployment_df)`. It describes the date, the numbers of senators, governors, and representatives in each of the two parties democrat or repliacn (gop), as well as whether the president is democrat or replican, the snp closing numbers, and the unemployment numbers. The years range from `r min(pull(pols_snp_unemployment_df, year))` to `r max(pull(pols_snp_unemployment_df, year))`. 

## Problem 2

Mr. Trash Wheel!

```{r q2_mr_trash_wheel}
mr_trash_wheel_df <- readxl::read_excel("./data/202309 Trash Wheel Collection Data.xlsx", 
                                        sheet = "Mr. Trash Wheel",
                                        range = "A2:N586") %>% 
  janitor::clean_names() %>% 
# The homework now says to omit rows that do not include dumpster specific data, but I think they all do?
  mutate(homes_powered = weight_tons*500/30) %>% 
  mutate(which_wheel = "Mr. Trash Wheel") %>%
  mutate(year = as.numeric(year))

prof_trash_wheel_df <- readxl::read_excel("./data/202309 Trash Wheel Collection Data.xlsx", 
                                        sheet = "Professor Trash Wheel",
                                        range = "A2:M108") %>% 
  janitor::clean_names() %>% 
# The homework now says to omit rows that do not include dumpster specific data, but I think they all do?
#  filter(is.na(x)) ?????????????
  mutate(homes_powered = weight_tons*500/30) %>% 
  mutate(which_wheel = "Professor Trash Wheel")
  
  # Note that Professor Trash Wheel does not have sports balls as a column

gwyn_trash_wheel_df <- readxl::read_excel("./data/202309 Trash Wheel Collection Data.xlsx", 
                                        sheet = "Gwynnda Trash Wheel",
                                        range = "A2:L157") %>% 
  janitor::clean_names() %>% 
# The homework now says to omit rows that do not include dumpster specific data, but I think they all do?
#  filter(is.na(x)) ?????????????
  mutate(homes_powered = weight_tons*500/30) %>% 
  mutate(which_wheel = "Gwynnda Trash Wheel")
  
```

Combine datasets into a single, tidy dataset 

```{r q2_combine}
# Not sure how to do this because some columns are missing 

trash_wheel_combined_df <- 
  bind_rows(mr_trash_wheel_df, prof_trash_wheel_df, gwyn_trash_wheel_df)

```

<PARAGRAPH ABOUT MR. TRASH WHEEL>
Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables. For available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of cigarette butts collected by Gwynnda in July of 2021?




## Question 3

Alzheimer's disease (AD) biomarkers and MCI 

#### Import, clean, and tidy the baseline demographics dataset, then describe it.

```{r q3_baseline_dataset}

mci_baseline_df <- read_csv("./data/data_mci/MCI_baseline.csv", skip = 1) %>% # The first column contains descriptions, so needs to be skipped to enable the column names in the df to be correct
  janitor::clean_names() %>% 
  mutate(sex = case_match(sex,
                          1 ~ "male",
                          0 ~ "female"
                          )) %>% 
  mutate(apoe4 = case_match(apoe4,
                            1 ~ "carrier",
                            0 ~ "noncarrier"
                            )) %>% 
  filter(!(age_at_onset <= current_age) | age_at_onset == ".") # Filter out subjects who had MCI at baseline

# Number of participants recruited who meet inclusion criteria

num_participants <- mci_baseline_df %>% 
  filter(age_at_onset > current_age) %>% 
  summarise(n())

# Number of participants who develop MCI

num_mci <- mci_baseline_df %>% 
  filter(age_at_onset > current_age) %>% 
  summarise(n())

# Proportion of women who are APOE4 carriers

prop_women_apoe4 <- mci_baseline_df %>% 
  filter(sex == "female") %>% 
  summarise(mean(apoe4 == "carrier"))

# Alternative method for future reference 
female_total <- nrow(mci_baseline_df %>% 
  filter(sex == "female"))
female_carrier <- nrow(mci_baseline_df %>% 
  filter(sex == "female") %>%
  filter(apoe4 == "carrier"))
prop_women_apoe4_bruteforce <- female_carrier/female_total

```

Importing baseline data: The original CSV had the first column as the description of the data, so needed to be skipped when imported. The column names were standardized with janitor::cleannames() for easier analysis. Additionally, sex and carrier status were coded using 0 and 1, so that was translated into words (male/female and carrier/noncarrier). Finally, the dataset contained subjects who had MCI at baseline, which is not what we are interested in for this analysis (they do not meet inclusion criteria), so they were filtered out. 

Features of baseline data: After filtering, `r num_participants` subjects met the inclusion criterion of no MCI at baseline, of which `r num_mci` developed MCI. The average baseline age was `r mci_baseline_df %>% summarise(mean(current_age))` years. The proportion of women in the study who are APOE4 carriers is `r prop_women_apoe4` (or `r prop_women_apoe4*100`%).

#### Import, clean, and tidy the longitudinally observed biomarkers dataset, then describe it. 

The following dataset contains measures of amyloid 42/40 ratios over 5 timepoints.

```{r q3_longitudinal_dataset}

mci_longitudinal_df <- read_csv("./data/data_mci/mci_amyloid.csv", skip = 1) %>%
  janitor::clean_names() %>% 
  pivot_longer("baseline":"time_8",
               names_to = "timepoint",
               values_to = "ratio")

### MAKE SOME COMMENTS

# In order to explore the overlap of subjects who appear in both datasets, I will use a wide format and inner_join. The longitudinal dataset is in wide format when importing, so I will import it, clean it, but not tidy it, and change the name of the column for study id to match the baseline dataset.

mci_longitudinal_df_wide <- read_csv("./data/data_mci/mci_amyloid.csv", skip = 1) %>%
  janitor::clean_names() %>% 
  rename(id = study_id)

# merge the two dfs
mci_merged_df <- inner_join(mci_baseline_df, mci_longitudinal_df_wide, by = "id")

# count the number of subjects that did not overlap
non_overlap <- anti_join(mci_baseline_df, mci_longitudinal_df_wide, by = "id") %>%
  summarise(n())

# Create a tidy version of the merged data
mci_merged_df_tidy <- mci_merged_df %>% 
    pivot_longer("baseline":"time_8",
               names_to = "timepoint",
               values_to = "ratio")

# Save merged, tidy dataset
write_csv(mci_merged_df_tidy, "./data/data_mci/mci_merged.csv")

```

Data import and merging: The longitudinal csv is in a wide format, so it was tidyed. However, in order to combine the longitudinal with the baseline df, the wide format was needed and the study_id/id column had to have the same name. Once they were compatible, they were merged over the common subjects using inner_join. The total number of subjects in the baseline df was `r mci_baseline_df %>% summarise(n())`, the total number in the longitudinal df were `r mci_longitudinal_df_wide %>% summarise(n())`, and the total number (overlapping) in the merged df was `r mci_merged_df %>% summarise(n())`. Although `r non_overlap` subjects did not appear in both datasets, the majority overlapped. 

Exported dataset csv: The exported, merged, tidied dataset contains the information from both datasets. It describes `r colnames(mci_merged_df_tidy)`. 
